Towards a Mechanized Proof of Noninterference

### 2013-12-19 paper/paper.tex

In this paper we discuss our experience trying to mechanize a proof of
noninterference for a simple language with security labels, \langName.  This
marks the first step towards writing a verified compiler for a language with
information flow types.  Although we did not complete the full noninterference
proof because we ran out of time, we were able to find small inaccuracies and
subtleties in the proofs from the paper we based our proof on.  We discuss what
would be involved in completing the proof and how this relates to other work.

@@

Verified systems have been growing in popularity recently with the rise of proof
assistants like Coq and NuPRL~\cite{coqart,nuprl}.  These systems allow a
programmer to write programs and have proofs, which are checked and compiled by
the system, which can assert various properties about the programs.  As we will
discuss in Section~\ref{sec:related}, these systems have been used to formalize,
prove correct, and implement systems that have verified properties.  This allows
for programmers to be more confident in their languages and programs, ensuring
that the properties they want are actually properties of their work.

Interestingly, as far as we know there has not been work to formalize and verify
a language, like Jif~\cite{myers1999jflow} or FlowCaml~\cite{simonet2003flow},
that has a type system for enforcing information flow security policies.  For
languages like these and other systems related to security, it is desirable for
the program to be accompanied with a proof of the security properties it
enforces.  Our goal is to make progress towards verifying the correctness of a
compiler for a language with information flow types that programmers will want
to use.

A verified compiler for a language like Jif would require a proof of
noninterference, which asserts that any program in the language will never
produce different outputs for low security items if only the high security input
values are varied.  Thus, in this paper we discuss our experience working
towards proving this theorem for a simple information flow type system as a
first step towards creating a verified information flow compiler.

Our language, \langName,  and proof are a simplified version of what is
presented in Simonet and Pottier's work on information flow for an ML-like
language, \origLang~\cite{InfoFlowML}.  This proof proceeds by designing a
language that includes values which represent two different alternatives,
representing the different values of high security items in two different
executions of the program.  The proof then becomes a matter of proving that if
the program is typed with a low security label, this does not produce a value
that could be different between two different executions of the program.

In this paper we will present our simplified language, \langName,
along with the semantics and type system (Section~\ref{sec:lang}).  We will then
present a natural language, human readable, proof of noninterference for the
language and discuss our experiences with mechanizing this proof in the Coq
proof assistant (Sections~\ref{sec:proof} and \ref{sec:mechanizing}).  Finally,
we will discuss related work, next steps, and conclude
(Sections~\ref{sec:related}, \ref{sec:future}, and \ref{sec:conclusion}).

@@

Our toy language for proving noninterference, \langName, is a simplified version
of \origLang from Simonet and Pottier's paper~\cite{InfoFlowML}.  Since we had a
limited amount of time to work on this project, we limited ourselves to a subset
of the language that drops the following features:

\begin{itemize}
  \item References and associated memory operations.
  \item Paired values.
  \item Exceptions.
  \item Sequencing operations.
\end{itemize}

While the language becomes much much less interesting for actual programming, we
feel that these features do not fundamentally change the problem of proving
noninterference.  In order to mechanize a proof of noninterference for a
language closer to \origLang, it would require simply adding mechanized proofs
for the additional cases where appropriate.

>>>>

In this paper we discuss our experience trying to mechanize a proof of
noninterference for a simple language with security labels, \langName, based on Pottier and Sionet's \origLang.  This
marks the first step towards writing a verified compiler for a language with
information flow types.  Although we did not complete the full noninterference
proof, we were able to find subtle inaccuracies in the \origLang proofs.  
We discuss what would be involved in completing the proof and how this 
relates to other work.

@@

Verified systems have been growing in popularity recently with the rise of proof
assistants like Coq and NuPRL~\cite{coqart,nuprl}.  
These systems allow programmers to construct proofs of properties of their programs which are checked and compiled by the system.  As we will
discuss in Section~\ref{sec:related}, these \emph{Proof Assistants} have been used to formalize,
prove correct, and implement systems with verified properties.  This increases Programmers' confidence in their languages and programs; to prove correctness of a function is a much stronger guarantee than simply testing functionality (or asserting correctness in the absence of tests).

The last decade has also seen a surge of research in languages which inforce information flow security, such as Jif~\cite{myers1999jflow} or FlowCaml~\cite{simonet2003flow}.  These languages provide features which programmers can use to enforce integrity and confidentiality of data as it is used throughout computations.

Given the growing popularity of Proof Assistants information-flow languages, we were surprised to discover 
that there has not been work done to formalize ad verify a language with a type systems which enforces information flow security policies.
These languages seem a natural choice for formal verification, as it is desirable for programs written in these languages to be accompanied with a proof of the security properties it enforces.
Our goal is to make progress towards verifying the correctness of a
compiler for a language with information flow types that programmers will want
to use.

The core theorem which distinguishes an information-flow language from a 
``normal'' programming language is the poof of noninterference, 
which asserts that across any two runs, every program in the language will never
produce different outputs for low security items if only the high security input
values are varied.  It is just this theorem we wish to prove in this paper; we
hope to learn through this effort of the unique challenges associated with
mechanizing proofs of information-flow languages.

Our language, \langName,  and proof are a simplified version of what is
presented in Simonet and Pottier's work on information flow for an ML-like
language, \origLang~\cite{InfoFlowML}.  The proof proceeds by designing a
core language, extending it with a term that represents a ``pair'' of 
values varied across two executions, and introducing these ``pairs'' of 
different values in context where the value would have a high-security label.
The proof then becomes a matter of showing that if
the program is typed with a low security label, it does not terminate
with a result that is an instance of one of these ``pairs.''

In this paper we will present our simplified language, \langName,
along with its semantics and type system (Section~\ref{sec:lang}).  We will then
present a natural language proof of noninterference for the
language and discuss our experiences with mechanizing this proof in the Coq
proof assistant (Sections~\ref{sec:proof} and \ref{sec:mechanizing}).  Finally,
we will discuss related work, next steps, and conclude
(Sections~\ref{sec:related}, \ref{sec:future}, and \ref{sec:conclusion}).

@@

Our toy language for proving noninterference, \langName, is a simplified version
of \origLang from Simonet and Pottier's paper~\cite{InfoFlowML}.  Since we had a
limited amount of time to work on this project, we restricted ourselves to a subset
of the language that drops the following features:

\begin{itemize}
  \item References and associated memory operations.
  \item Value Tuples.
  \item Exceptions.
  \item Sequencing operations.
\end{itemize}

While the language becomes much more burdensome for the programmer, we
feel that these features do not fundamentally change the problem of proving
noninterference.  Mechanizing a proof of noninterference for a
language closer to \origLang would require simply adding mechanized proofs
for the additional cases where appropriate.
